Dynastic Extinctions and Urban Economic Development: \\Evidence from the Holy Roman Empire\footnote{I thank Davide Cantoni, Matthias Weigand, and my RA colleagues at the Chair of Economic History for helpful comments and feedback.}

### 2023-06-09 paper/paper.tex
#22 add a lot more literature to intro

\title{Dynastic Extinctions and Urban Economic Development: \\Evidence from the Holy Roman Empire\footnote{
I thank Davide Cantoni, Matthias Weigand, and my RA colleagues at the Chair of Economic History for helpful comments and feedback. I thank Matthias Weigand in particular for providing me with his own data and for pointing me towards the paper by \cite{schoenholzer2022}.
}
}
\author{Elias Hadj Ammar}
\date{May 2023}

@@

The history of Europe was defined by conflict. No other continent at any time experienced more war than Europe in the early modern era (\citealp{voigtlnder2013}, p. 174). Wars brought death and devastation to the societies affected by them, but their effects also manifested on the political map --- they play out in a diverse ecosystem of polities that come and go and change size over time as the borders get redrawn and territory changes hands. One can indeed imagine this as an evolutionary process: states compete for resources and powerful states eventually absorb weaker ones. As a result, traits that increase state power --- the ability to win conflicts against rivals --- get selected for, while other traits are doomed to disappear \citep{levine2013, levine2021}.

In this paper I study the role of this process in the economic development of Europe. In particular, I estimate the effect of state power, the trait that conflict selects for, on urban development. If the effect is positive, then selection of "better" states through conflict-driven evolution may partially explain the ascendancy of Europe.

Past treatments of this topic have suffered from a lack of available historical data. \cite{diamond1997} and \cite{landes2006} each present narrative arguments for the economic benefits of state competition but stop short of formal empirical analysis. Such a study requires at the very least three things: 1) a detailed territorial history of European cities, 2) some city-level measure of development reaching far enough into the past, 3) some measure of state power, institutional quality, or whatever else one wishes to explain effects with.
These data limitations are starting to be overcome. Recently \cite{schoenholzer2022} have investigated how switching between states affects the development of cities. They find that switching states affects population negatively in the short run but positively in the long run. Similar to the evolution metaphor, they interpret this as "creative destruction": the short run cost of destruction is compensated by the long-run benefit of becoming part of a "better" state. However, their

The analysis in this paper extends that of \cite{schoenholzer2022} by identifying special cases of switching events: in \textit{conquests} a city is forcefully absorbed into a new state, while in \textit{successions} a city is absorbed into a new state after the rulers of the old state have gone extinct. Comparing the effects of these two types of switches allows me to tease out the effect of state power.

I make a key assumption: conquests imply state power differentials but successions do not. In other words, if we consider the universe of all switching events, the average increase in state power from the old to the new owner would be higher for the conquests than for the successions. The reasoning is that conflict --- as in a conquest --- selects the new state for power, but conflict-free successions, depending mainly on dynastic marriages, are more "random". This assumption allows me to use conquests as a stand-in for an increase in state power.

@@

\cite{diamond1997} asserts that the geographical remoteness of China deprived it of the benefits of state competition that European states enjoyed throughout their history. \cite{landes2006} makes a similar argument, pointing out how the totalitarianism and bureaucratism that stifled innovation in medieval China would not have survived long in a more fragmented state system.

@@

>>>>

\title{Conflict Driven Evolution and Urban Development: \\ Evidence from the Holy Roman Empire\footnote{
I thank Davide Cantoni, Matthias Weigand, and my RA colleagues at the Chair of Economic History for helpful comments and feedback. I thank Matthias Weigand in particular for providing me with his own data and for pointing me towards the paper by \cite{schoenholzer2022}.
}
}
\author{Elias Hadj Ammar}
\date{May 2023}

@@

The European state system was forged in conflict. No other continent at any time experienced more war than Europe in the early modern era (\citealp{voigtlnder2013}, p. 174). Wars brought death and devastation to the societies affected by them, but their effects also manifested on the political map --- they play out in a diverse ecosystem of polities that come and go and change size over time. Borders get redrawn and territory changes hands. One can indeed imagine this as an evolutionary process: states compete for resources and powerful states eventually absorb weaker ones. As a result, traits that increase state power --- the ability to win conflicts against rivals --- get selected for, while unfavourable traits are doomed to disappear \citep{levine2013, levine2021}.

In this paper I study the role of this process in the economic development of Europe. In particular, I estimate the effect of state power, the trait that conflict selects for, on urban development. If the effect is positive, then the conflict driven evolution of states may partially explain the eventual ascendancy of Europe: by selecting for military success, Europe's violent history may have accidentally produced fiscal capacity, efficient legal systems, and prospering middle classes.

The idea that war and conflict can raise the standard of living dates back to \cite{malthus1798}. He reasoned that "positive checks" on the population (war, disease, famine) lead to higher land-labour ratios and therefore higher levels of income per capita. \cite{voigtlnder2013} build upon this Malthusian mechanism and add an explanation for the incessant wars: taxes on those higher incomes filled the treasuries of the rulers, which increased their demand for war --- a luxury good ---, which in turn kept population low.

My story is not about population but institutions (although it is compatible with Malthusianism). Explaining long-run growth with institutional factors has a long tradition in economics: North, Acemoglu, property rights, civil law, middle class. The key institutional change brought about by conflict was fiscal capacity \citep{tilly1985, dincecco2012}, or the state's ability to raise taxes. A newer strand of research focuses on fiscal capacity:  \cite{cantoni2023}

Theory of wars and states: \cite{tilly1985} argues that war was the main reason for governments to invest in fiscal capacity. The formal theory of conflict-driven evolution is from \cite{levine2013}.

\cite{diamond1997} and \cite{landes1969, landes2006} have argued for the economic benefits of political fragmentation and state competition. \cite{diamond1997} asserts that geography 
remoteness of China deprived it of the benefits of state competition that European states enjoyed throughout their history. \cite{landes2006} makes a similar argument, pointing out how the totalitarianism and bureaucratism that stifled the spread of innovation in medieval China would not have survived long in a more fragmented state system.

The closest relative to this paper is \cite{schoenholzer2022} who investigate how switching between states affects the development of cities. They find that switching states affects population negatively in the short run but positively in the long run. Similar to the evolution metaphor, they interpret this as "creative destruction": the short run cost of destruction is compensated by the long-run benefit of becoming part of a higher-quality state. However, they do not show explicitly how these improvements in state quality are driven by selection through conflict.

\hspace{0.5cm}

The analysis in this paper builds upon that of \cite{schoenholzer2022} and extends it by identifying special cases of switching events: in \textit{conquests} a city is forcefully absorbed into a new state, while in \textit{successions} a city is absorbed into a new state after the rulers of the old state have gone extinct. Comparing the effects of these two types of switches allows me to tease out the effect of state power. I find a positive and significant effect of state power on the long-run development of cities. This result suggests that conflict-driven evolution indeed played a role in the rise of Europe relative to other regions of the world.

I make a key assumption: conquests imply state power differentials but successions do not. Specifically, if we consider the universe of all switching events, the average increase in state power from the old to the new owner would be higher for the conquests than for the successions. The reasoning is that conflict --- as in a conquest --- selects the new state for power, but conflict-free successions, depending mainly on dynastic marriages, are more "random". This assumption allows me to use conquests as a stand-in for an increase in state power. Several other assumptions are needed to interpret my results causally, which I will discuss in detail below.

@@

@@

\section{Data}

Empirical tests have suffered from a lack of available historical data. Such a study requires at the very least three things: 1) a detailed territorial history of European cities, 2) some city-level measure of development reaching far enough into the past, and 3) some measure of state power, institutional quality, or whatever else one wishes to explain effects with. These data limitations are starting to be overcome.
